New evidence of self-care tooling and phylogenetic modelling reveal parrot tool use is not rare

### 2025-02-05 manuscript.tex
revise and resubmit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man, donotrepeattitle,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{tool use; technical intelligence; parrots; digital video platforms; phylogenetic modelling}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright † These authors contributed equally to this work \par ‡ These authors contributed equally to this work \par * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Evidence of self-care tooling on digital video platforms and phylogenetic modelling reveal parrot tool use is not rare},
  pdfauthor={Amalia P. M. Bastos†,1,2, Scott Claessens*†,2, Ximena J. Nelson3, David Welch4, Quentin D. Atkinson‡,2, \& Alex H. Taylor‡,2,3,5,6},
  pdflang={en-EN},
  pdfkeywords={tool use; technical intelligence; parrots; digital video platforms; phylogenetic modelling},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Evidence of self-care tooling on digital video platforms and phylogenetic modelling reveal parrot tool use is not rare}
\author{Amalia P. M. Bastos\textsuperscript{†,1,2}, Scott Claessens\textsuperscript{*†,2}, Ximena J. Nelson\textsuperscript{3}, David Welch\textsuperscript{4}, Quentin D. Atkinson\textsuperscript{‡,2}, \& Alex H. Taylor\textsuperscript{‡,2,3,5,6}}
\date{}

@@

Comparative work on the evolution of tool use is a paradigmatic example of this
issue. The initial discoveries of tool use in chimpanzees\textsuperscript{6},
birds\textsuperscript{7}, dolphins\textsuperscript{8}, and octopuses\textsuperscript{9} occurred
decades after significant advances on other more easily measurable aspects of
their biology. Since then, scholars have proposed a clear operational
definition of tool use applicable to all species --- the manipulation of an
unattached object as an extension of the animal's body to achieve a
goal\textsuperscript{10} --- and have used the distribution of species meeting this
definition to make various claims about the evolutionary drivers of tool use
behaviours. For example, based on the observation that bird species with
reported tool use tend to have larger brains, researchers have identified
higher relative brain size as a likely precondition for tool using
capabilities\textsuperscript{11--14}. These
researchers argue that larger brains are better able to integrate visual and
somatosensory information when innovating novel behaviours, such as tool use,
in changing environments\textsuperscript{15,16} (but see ref\textsuperscript{17}).
Similarly, researchers have used existing reports of tool use in birds to debate
the roles of generalist versus specialist feeding strategies in driving the
evolution of tool use, with some arguing that feeding generalists require
technical innovations to expand their dietary niche\textsuperscript{15,18,19} and others arguing that feeding
specialists require technical innovations for extractive foraging of specific
foods\textsuperscript{20,21}.

However, before we can make claims of this kind, we need to know whether current
research effort in the literature is sufficient for robust conclusions to be
drawn about the evolution of tool use. In fact, evidence suggests that research
effort is often systematically biased towards particular taxonomic groups, parts
of the world that are easy to access, and species with life history traits that
make them easier to study, such as larger distribution ranges and population
sizes\textsuperscript{22}. This is a crucial limitation because insufficient
observation may lead researchers to miss true instances of tool behaviours and
thus draw premature conclusions about the evolutionary drivers and origins of
tool use. Researchers are keenly aware of this problem and have attempted to
deal with it in different ways, such as controlling for the number of scientific
papers published on different species. But previous work has not yet attempted
to quantify and explicitly model the relationship between actual tool-using
behaviour and what is reported in the scientific literature. If more tool-using
species exist than previously thought, this could have important implications
for theories of the evolutionary drivers and origins of tool use and for our
understanding of how rare this behaviour actually is.

@@

Figure \ref{fig:plotPhylo1} maps the findings from the video survey onto a
maximum clade credibility phylogeny for the parrot order, plotted alongside
species previously identified in the scientific literature. Before the video
survey, 11 parrot species (3\%) had been identified as tool users in the
scientific literature. Across our video survey, we observed true tool use for self-care in
22
species, 5 of which overlapped with the scientific literature and
17
of which were novel species. All of the species identified in the video survey
were cockatoos (\emph{Cacatuidae}), Old World parrots (\emph{Psittacinae}), or neotropical
parrots (\emph{Arinae}). The most common species in our survey, accounting for
41 videos from
37
individuals, was the green-cheeked conure (\emph{Pyrrhura molinae}). In accordance
with the scientific literature, the video survey found no evidence of tool use
in any species of Psittaculidae, despite this family containing some of the most
commonly kept pet species, including lovebirds, lorikeets, and Asian parakeet
species. Combining both the video survey and the scientific literature, we can
thus identify 28 total parrot species (7\%) capable of true tool use as defined
by ref\textsuperscript{10}, compared to the 11 previously reported.

@@

Our data raise the question of why such a wide range of parrot species exhibit
self-scratching tool use behaviours in captive environments. It is possible that
these self-care behaviours emerge more often in captivity because pet parrots
are kept with few or no conspecifics, and therefore must innovate tool use to
preen inaccessible body areas, such as their heads and necks, which would have
been otherwise preened by flock members. This may indicate that human
interaction and allopreening in the absence of a conspecific is insufficient for
pet parrots, as suggested by studies on welfare and chronic stress in lone
captive parrots\textsuperscript{63--66}. Beyond
social isolation in captivity, other factors that reduce preening efficacy by an
individual might also lead to the innovation of self-care tooling: for example,
a captive kea housed with conspecifics in a naturalistic environment was shown
to have innovated the use of a pebble tool for self-care likely as a consequence
of a missing upper mandible\textsuperscript{31}. These behaviours might therefore be
innovated by individuals when ecologically necessary. Critically, however, a
capacity to innovate a behaviour such as tool use must exist in a species in
order to manifest in any environment, whether that be captive or in the wild.
The fact that only a subset of parrot species are innovating self-care tooling
in response to their captive setting and that self-care tooling is predicted by
phylogenetic history and encephalisation strongly suggests that self-scratching
indicates a capacity to produce tool use, which is expressed in different ways
in captive and wild settings.

Our findings have a number of important implications. First, our findings show
that current research effort in the scientific literature is insufficient to
capture the real world occurrence of parrot tool use. If the scientific
literature had sampled the natural world sufficiently, we would expect to see
close correspondence between those species reported as tool users in the
literature and those species the public have filmed performing tool use.
Instead, we discovered a large discrepancy between these two data sources, both
in the prevalence of tool use and the species identified, likely due to the
numerous difficulties inherent to conducting observations in the wild. This
raises the possibility that other rare behaviours remain yet to be discovered in
the scientific literature.

@@

Third, our results support existing theories of the drivers of tool use. We
found that encephalisation was strongly positively related to the probability of
tool use in our phylogenetic model, supporting previous theories linking
relative brain size to increased tool innovation in birds\textsuperscript{12--14} and primates\textsuperscript{67}. We acknowledge that
encephalisation quotient is not a perfect measure due to measurement error and
challenges with interpretation\textsuperscript{3,4}. However, encephalisation
quotient has much greater coverage across the parrot phylogeny than more
fine-grained measures like whole neuron count\textsuperscript{68} and there is no reason
to think that measurement error would produce the consistent patterns across our
study and prior work. To understand the causal mechanisms responsible for these
relationships, we encourage further work on the specific neural correlates of
technical intelligence in parrots, e.g.,\textsuperscript{69}. In our phylogenetic
model, we also found that tool use was somewhat more likely among feeding
generalists compared to feeding specialists, although this difference was
uncertain. This trend supports previous suggestions that increased cognitive
abilities and technical innovation rates are required to expand a generalist
species' dietary niche\textsuperscript{18,20,70}. However, the trend contradicts theories linking tool use to
dietary specialisation, whereby species eating specific foods that require
extractive foraging have higher cognitive ability and are especially prone to
using tools\textsuperscript{20}.

A potential concern with data originating from digital video platforms is its
reliability\textsuperscript{23,71--75}.
In particular, the self-care tool use instances detected in our video survey
might be attributable to training or unintentional cueing by the birds' owners.
However, there was little evidence to suggest that the observations of
self-scratching tool use in the video survey were merely unintentional accidents
or explicitly trained behaviours. Individual parrots often used tools slowly and
repetitively over long periods of time, even across multiple different videos,
suggesting that their behaviour was not random or accidental\textsuperscript{74}.
Parrots preferentially employed self-scratching tools on areas of their body
that were otherwise inaccessible, with 96\% of all instances involving scratching
of the head or neck, suggesting intentional tool use. For 60\% of the species in
the video survey, we found two or more videos of repetitive and sustained
scratching by different individuals of the same species in separate households,
suggesting that the manipulations were intentional and recurring events that did
not represent unusual stereotypies of any single individual. Regarding the
possibility of training or cueing from owners, over half of the videos contained
no evidence of human interaction aside from filming the behaviour. Humans only
handed parrots their tools in two of the videos, and none of the videos featured
owners directly rewarding tool use behaviours with food. It is unlikely that
owners would specifically train parrots to scratch themselves with tools.
Finally, the high levels of phylogenetic signal in our data provide strong
evidence that the observations from our video survey reflect
biologically-endowed capacities for tool use rather than accidental or trained
behaviours, which would likely appear uniformly across the phylogeny.
Nevertheless, future work using data such as this should choose target
behaviours carefully and examine videos for evidence of training or cueing.

While missing data imputation is becoming more common in phylogenetic
analyses\textsuperscript{76}, the important distinction between absence of
evidence and evidence of absence has not been given as much attention. Our
phylogenetic analysis provides one approach to this problem by distinguishing
between true absences of tool use and absences of tool use due to a lack of
research effort in the scientific literature or in videos. To achieve this, we
explicitly modelled the measurement of the outcome variable along a research
effort time series, such that species with lower research effort in the
literature or in videos were likely to be censored. In line with our causal
model, we also included relative brain size, feeding strategy, and phylogenetic
history as predictors of unobserved tool use. We encourage researchers to test
this model by directing future study efforts towards the parrot species with the
highest probabilities of being undetected tool-users. Future work should also
refine the causal model in Figure \ref{fig:plotDAG} to provide more certain
estimates of tool use probabilities, either by including additional predictor
variables or modelling further causes of measurement error in the taxonomic
record, such as species abundance and geographic accessibility\textsuperscript{22}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Auersperg2020}{}}%
\CSLLeftMargin{71. }%
\CSLRightInline{Auersperg, A.M.I., Schwing, R., Mioduszewska, B., O'Hara, M., and Huber, L. (2020). Do puffins use tools? Proceedings of the National Academy of Sciences \emph{117}, 11859--11859. \url{https://doi.org/10.1073/pnas.2001988117}.}

\leavevmode\vadjust pre{\hypertarget{ref-DechaumeMoncharmont2020}{}}%
\CSLLeftMargin{72. }%
\CSLRightInline{Dechaume-Moncharmont, F.-X. (2020). Touchy matter: The delicate balance between {M}organ's canon and open-minded description of advanced cognitive skills in the animal. Peer Community in Ecology, 100042. \url{https://doi.org/10.24072/pci.ecology.100042}.}

\leavevmode\vadjust pre{\hypertarget{ref-Farrar2020}{}}%
\CSLLeftMargin{73. }%
\CSLRightInline{Farrar, B.G. (2020). Evidence of tool use in a seabird? PsyArXiv. \url{https://doi.org/10.31234/osf.io/463hk}.}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man, donotrepeattitle,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\makeatletter
\usepackage{etoolbox}
\patchcmd{\maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{tool use; technical intelligence; parrots; digital video platforms; phylogenetic modelling}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright † These authors contributed equally to this work \par ‡ These authors contributed equally to this work \par * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={New evidence of self-care tooling and phylogenetic modelling reveal parrot tool use is not rare},
  pdfauthor={Amalia P. M. Bastos†,1,2, Scott Claessens*†,2, Ximena J. Nelson3, David Welch4, Quentin D. Atkinson‡,2, \& Alex H. Taylor‡,2,3,5,6},
  pdflang={en-EN},
  pdfkeywords={tool use; technical intelligence; parrots; digital video platforms; phylogenetic modelling},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{New evidence of self-care tooling and phylogenetic modelling reveal parrot tool use is not rare}
\author{Amalia P. M. Bastos\textsuperscript{†,1,2}, Scott Claessens\textsuperscript{*†,2}, Ximena J. Nelson\textsuperscript{3}, David Welch\textsuperscript{4}, Quentin D. Atkinson\textsuperscript{‡,2}, \& Alex H. Taylor\textsuperscript{‡,2,3,5,6}}
\date{}

@@

Comparative work on the evolution of tool use is a paradigmatic example of this
issue. The initial discoveries of tool use in chimpanzees\textsuperscript{6},
birds\textsuperscript{7}, dolphins\textsuperscript{8}, and octopuses\textsuperscript{9} occurred
decades after significant advances on other more easily measurable aspects of
their biology. Since then, scholars have proposed a clear operational
definition of tool use applicable to all species --- the manipulation of an
unattached object as an extension of the animal's body to achieve a
goal\textsuperscript{10} --- and have used the distribution of species meeting this
definition to make various claims about the evolutionary drivers of tool use
behaviours. For example, based on the observation that bird species with
reported tool use tend to have larger brains, researchers have identified
higher relative brain size as a likely precondition for tool using
capabilities\textsuperscript{11--14}. These
researchers argue that larger brains are better able to integrate visual and
somatosensory information when innovating novel behaviours, such as tool use,
in changing environments\textsuperscript{15,16} (but see Kaplan\textsuperscript{17}).
Similarly, researchers have used existing reports of tool use in birds to debate
the roles of generalist versus specialist feeding strategies in driving the
evolution of tool use, with some arguing that feeding generalists require
technical innovations to expand their dietary niche\textsuperscript{15,18,19} and others arguing that feeding
specialists require technical innovations for extractive foraging of specific
foods\textsuperscript{20,21}.

However, before we can make claims of this kind, we need to know whether current
research effort in the literature is sufficient for robust conclusions to be
drawn about the evolution of tool use. In fact, evidence suggests that research
effort is often systematically biased towards particular taxonomic groups, parts
of the world that are easy to access, and species with life history traits that
make them easier to study, such as larger distribution ranges and population
sizes\textsuperscript{22}. This is a crucial limitation because insufficient
observation may lead researchers to miss true instances of tool behaviours and
thus draw premature conclusions about the evolutionary drivers and origins of
tool use. Researchers are keenly aware of this problem and have attempted to
deal with it in different ways, such as controlling for the number of scientific
papers published on different species. However, previous work has not yet
attempted to quantify and explicitly model the relationship between actual
tool-using behaviour and what is reported in the scientific literature. If more
tool-using species exist than previously thought, this could have important
implications for theories of the evolutionary drivers and origins of tool use
and for our understanding of how rare this behaviour actually is.

@@

Figure \ref{fig:plotPhylo1} maps the findings from the video survey onto a
maximum clade credibility phylogeny for the parrot order, plotted alongside
species previously identified in the scientific literature. Before the video
survey, 11 parrot species (3\%) had been identified as tool users in the
scientific literature. Across our video survey, we observed true tool use for self-care in
22
species, 5 of which overlapped with the scientific literature and
17
of which were novel species. All of the species identified in the video survey
were cockatoos (\emph{Cacatuidae}), Old World parrots (\emph{Psittacinae}), or neotropical
parrots (\emph{Arinae}). The most common species in our survey, accounting for
41 videos from
37
individuals, was the green-cheeked conure (\emph{Pyrrhura molinae}). In accordance
with the scientific literature, the video survey found no evidence of tool use
in any species of Psittaculidae, despite this family containing some of the most
commonly kept pet species, including lovebirds, lorikeets, and Asian parakeet
species. Combining both the video survey and the scientific literature, we can
thus identify 28 total parrot species (7\%) capable of true tool use as defined
by Shumaker et al.\textsuperscript{10}, compared to the 11 previously reported.

@@

Our data raise the question of why such a wide range of parrot species exhibit
self-scratching tool use behaviours in captive environments. It is possible that
these self-care behaviours emerge in captivity because pet parrots are kept with
few or no conspecifics, and therefore must innovate tool use to preen
inaccessible body areas, such as their heads and necks, which would have been
otherwise preened by flock members. This may indicate that human interaction and
allopreening in the absence of a conspecific is insufficient for pet parrots, as
suggested by studies on welfare and chronic stress in lone captive parrots\textsuperscript{63--66}. Beyond social
isolation in captivity, other factors that reduce preening efficacy by an
individual might also lead to the innovation of self-care tooling: for example,
a captive kea housed with conspecifics in a naturalistic environment was shown
to have innovated the use of a pebble tool for self-care likely as a consequence
of a missing upper mandible\textsuperscript{31}. These behaviours might therefore be
innovated by individuals when ecologically necessary. Critically, however, a
capacity to innovate a behaviour such as tool use must exist in a species in
order to manifest in any environment, whether that be captive or in the wild.
The fact that only a subset of parrot species are innovating self-care tooling
in response to their captive setting and that self-care tooling is predicted by
phylogenetic history and encephalisation strongly suggests that self-scratching
indicates a capacity to produce tool use, which may be expressed in different
ways in captive and wild settings. Interestingly, at present there is no
evidence to suggest that parrots exhibit self-scratching tool use in the wild.
This could reflect a true absence of this behaviour in wild parrots, which may
not experience the same environmental pressures that lead to the emergence of
self-scratching in captivity. On the other hand, it could also reflect a false
negative, with self-scratching occurring sufficiently infrequently or
inconspicuously in wild populations that it remains unobserved or unreported.

Our findings have a number of important implications. First, our findings
suggest that current research effort in the scientific literature is
insufficient to capture the real world occurrence of parrot tool use. If the
scientific literature had sampled the natural world sufficiently, we would
expect to see close correspondence between those species reported as tool users
in the literature and those species the public have filmed performing tool use.
Instead, we discovered a large discrepancy between these two data sources, both
in the prevalence of tool use and the species identified, likely due to the
numerous difficulties inherent to conducting observations in the wild. This
raises the possibility that other rare behaviours remain yet to be discovered in
the scientific literature.

@@

Third, our results support existing theories of the drivers of tool use. We
found that encephalisation was strongly positively related to the probability of
tool use in our phylogenetic model, supporting previous theories linking
relative brain size to increased tool innovation in birds\textsuperscript{12--14} and primates\textsuperscript{67}. We acknowledge that
encephalisation quotient is not a perfect measure due to measurement error and
challenges with interpretation\textsuperscript{3,4}. However, encephalisation
quotient has much greater coverage across the parrot phylogeny than more
fine-grained measures like whole neuron count\textsuperscript{68} and there is no reason
to think that measurement error would produce the consistent patterns across our
study and prior work. To understand the causal mechanisms responsible for these
relationships, we encourage further work on the specific neural correlates of
technical intelligence in parrots, e.g., Cabrera-Álvarez and Clayton\textsuperscript{69}. In our phylogenetic model, we also found that tool use was
somewhat more likely among feeding generalists compared to feeding specialists,
although this difference was uncertain. This trend supports previous suggestions
that increased cognitive abilities and technical innovation rates are required
to expand a generalist species' dietary niche\textsuperscript{18,20,70}. However, the trend contradicts theories
linking tool use to dietary specialisation, whereby species eating specific
foods that require extractive foraging have higher cognitive ability and are
especially prone to using tools\textsuperscript{20}.

While missing data imputation is becoming more common in phylogenetic
analyses\textsuperscript{71}, the important distinction between absence of
evidence and evidence of absence has not been given as much attention. Our
phylogenetic analysis provides one approach to this problem by distinguishing
between true absences of tool use and absences of tool use due to a lack of
research effort in the scientific literature or in videos. To achieve this, we
explicitly modelled the measurement of the outcome variable along a research
effort time series, such that species with lower research effort in the
literature or in videos were likely to be censored. In line with our causal
model, we also included relative brain size, feeding strategy, and phylogenetic
history as predictors of unobserved tool use. We encourage researchers to test
this model by directing future study efforts towards the parrot species with the
highest probabilities of being undetected tool-users. Future work should also
refine the causal model in Figure \ref{fig:plotDAG} to provide more certain
estimates of tool use probabilities, either by including additional predictor
variables or modelling further causes of measurement error in the taxonomic
record, such as species abundance and geographic accessibility\textsuperscript{22}.

\hypertarget{limitations-of-study}{%
\subsection{Limitations of study}\label{limitations-of-study}}

A potential concern with data originating from digital video platforms is its
reliability\textsuperscript{23,72--76}.
In particular, the self-care tool use instances detected in our video survey
might be attributable to training or unintentional cueing by the birds' owners.
However, there was little evidence to suggest that the observations of
self-scratching tool use in the video survey were merely unintentional accidents
or explicitly trained behaviours. Individual parrots often used tools slowly and
repetitively over long periods of time, even across multiple different videos,
suggesting that their behaviour was not random or accidental\textsuperscript{75}.
Parrots preferentially employed self-scratching tools on areas of their body
that were otherwise inaccessible, with 96\% of all instances involving scratching
of the head or neck, suggesting intentional tool use. For 60\% of the species in
the video survey, we found two or more videos of repetitive and sustained
scratching by different individuals of the same species in separate households,
suggesting that the manipulations were intentional and recurring events that did
not represent unusual stereotypies of any single individual. Regarding the
possibility of training or cueing from owners, over half of the videos contained
no evidence of human interaction aside from filming the behaviour. Humans only
handed parrots their tools in two of the videos, and none of the videos featured
owners directly rewarding tool use behaviours with food. It is unlikely that
a range of owners would independently train parrots to scratch themselves with
tools. Finally, the high levels of phylogenetic signal in our data provide
strong evidence that the observations from our video survey reflect
biologically-endowed capacities for tool use rather than accidental or trained
behaviours, which would likely appear uniformly across the phylogeny.
Nevertheless, future work using data such as this should choose target
behaviours carefully and examine videos for evidence of training or cueing.

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Debastiani2021}{}}%
\CSLLeftMargin{71. }%
\CSLRightInline{Debastiani, V.J., Bastazini, V.A.G., and Pillar, V.D. (2021). Using phylogenetic information to impute missing functional trait values in ecological databases. Ecological Informatics \emph{63}, 101315. \url{https://doi.org/10.1016/j.ecoinf.2021.101315}.}

\leavevmode\vadjust pre{\hypertarget{ref-Auersperg2020}{}}%
\CSLLeftMargin{72. }%
\CSLRightInline{Auersperg, A.M.I., Schwing, R., Mioduszewska, B., O'Hara, M., and Huber, L. (2020). Do puffins use tools? Proceedings of the National Academy of Sciences \emph{117}, 11859--11859. \url{https://doi.org/10.1073/pnas.2001988117}.}

\leavevmode\vadjust pre{\hypertarget{ref-DechaumeMoncharmont2020}{}}%
\CSLLeftMargin{73. }%
\CSLRightInline{Dechaume-Moncharmont, F.-X. (2020). Touchy matter: The delicate balance between {M}organ's canon and open-minded description of advanced cognitive skills in the animal. Peer Community in Ecology, 100042. \url{https://doi.org/10.24072/pci.ecology.100042}.}
